---
title: "Assignment 2: Coding Basics"
author: "Grace Randall"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. 
question_one <- seq(1,30,3)#forming a sequence from 1 to 30 with jumps of 3
question_one #checking output of this sequence
#2. 
mean_q_one <- mean(question_one) # calculating the mean of question_one
median_q_one <- median(question_one) # calculating the median of question_one
mean_q_one #printing the mean 
median_q_one #printing the median
#3. 
(mean_q_one>median_q_one)# checking if mean is greater than median 

```

## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r basics 2}
name <- c("Alice", "Bob", "Charlie", "Dan", "Eve") #character string vector of students
name <- as.data.frame(name)
score_v <- c(35,73,73,17,31) # number vector or scores
score <- as.data.frame(score_v)
did_they_pass <- c(FALSE,TRUE,TRUE,FALSE,FALSE) # Boolean vector 
did_they_pass <- as.data.frame(did_they_pass)
test_results <- cbind(name,score,did_they_pass)
class(test_results)
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: If this were held in a matrix, then all of the column's would need to be of the same data type. in this case they would all be turned into chr. 

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r basics 3}
pass_or_fail <- function(scores){ #create function
  
for(x in 1:length(scores)){ # loops for all of the values given in the vector
ifelse (scores[x]>=50, print(TRUE), print(FALSE)) #checks if passes of fails
  
}}
pass_or_fail(score_v) #testing function on answer from 5


```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: Both ifelse or if and else would work for this. Since there are only two options for what to do that can be defined by a single conditional ifelse is sufficient for this function. Anything that can be written using ifelse can also be written using if and else. 