---
title: "Assignment 5: Data Visualization"
author: "Grace Randall"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: inline
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1.  Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 
#bring in packages
library(tidyverse)
library(lubridate)
library(here)
library(cowplot)
library(ggthemes)

#load data
peterpaul_chem <- read.csv(
  here("Data","Processed_KEY",
       "NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"),
  stringsAsFactors = TRUE)
litter <- read.csv(
  here("Data","Processed_KEY",
       "NEON_NIWO_Litter_mass_trap_Processed.csv"),
  stringsAsFactors = TRUE)

#2 
#fix dates for lakes
class(peterpaul_chem$sampledate)
peterpaul_chem$sampledate <- ymd(peterpaul_chem$sampledate)

#fix months for lakes
peterpaul_chem$month <- 
  factor(
    peterpaul_chem$month,
    levels=c("1","2","3","4","5","6","7","8","9","10","11","12"),
                               labels = month.abb) 

#filter litter for needles
litter <- filter(litter, functionalGroup == "Needles")
#fix litter dates
litter$collectDate <- ymd(litter$collectDate)

```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3
my_theme <- theme_base()  + 
  theme(
    #line =              element_line(),
    #rect =              element_rect(),
    #text =              element_text(),
    
    # Modified inheritance structure of text element
    plot.title =        element_text(color = "midnightblue"),
    axis.title.x =      element_text(color = "midnightblue"),
    axis.title.y =      element_text(color = "midnightblue"),
    #axis.text =         element_text(),
    
    # Modified inheritance structure of line element
    #axis.ticks =        element_line(),
    panel.grid.major =  element_line(color="white"),
    #panel.grid.minor =  element_blank(),
    
    # Modified inheritance structure of rect element
    #plot.background =   element_rect(),
    panel.background =  element_rect(fill = "lightskyblue1"),
    #legend.key =        element_rect(),
    
    # Modifiying legend.position
    #legend.position = 'top',
    
    #complete = TRUE
    )

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 
ggplot(peterpaul_chem, aes(x=po4,y=tp_ug,color=lakename))+ 
  xlim(0,45)+ #get rid of extreme values
  labs(title = "Total Phosporus vs. Phosphate", color = "Lake")+ #set labels
  xlab("Phosphate")+
  ylab("Total Phosphorus")+
  geom_point()

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tip:
* Recall the discussion on factors in the previous section as it may be helpful here. 
* R has a built-in variable called `month.abb` that returns a list of months;see <https://r-lang.com/month-abb-in-r-with-example>

```{r Create boxplots}
#5 

#making temperature plot
 temp<-ggplot(peterpaul_chem)+
  theme(legend.position="none")+
  geom_boxplot(aes(x=month,y=temperature_C,color=lakename))+
  ylab("Temp (c)")+
  scale_x_discrete(drop=FALSE)
 
#making tp plot
 tp<-ggplot(peterpaul_chem)+
  theme(legend.position="none")+
  ylab("Total Phosphorus (um)")+
  geom_boxplot(aes(x=month,y=tp_ug,color=lakename))+
  scale_x_discrete(drop=FALSE)

 #making tn plot
 TN<-ggplot(peterpaul_chem)+
  theme(legend.position="none")+
  ylab("Total Nitrogen (um)")+
  geom_boxplot(aes(x=month,y=tn_ug,color=lakename))+
  scale_x_discrete(drop=FALSE)
 
 #grabing the legend
 shared_legend <-get_legend(TN + theme(legend.position = "bottom"))
 
 #plotting all the parts
 plot_grid(temp,tp,TN,shared_legend,nrow=4,align = 'h')

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: temperatures are clearly higher in both lakes in the summer months. there is not as clear of a seasonal trend in the nutrients but both Nitrogen and Phosphorus are higher in peter lake.

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6
ggplot(litter,aes(x=collectDate,y=dryMass,color=nlcdClass))+
  xlab("Collection Date")+
  ylab("Dry Mass")+
  theme(legend.position = "top")+
  geom_point()

#7
ggplot(litter,aes(x=collectDate,y=dryMass))+
  xlab("Collection Date")+
  ylab("Dry Mass")+
  my_theme+
  facet_wrap(vars(nlcdClass),nrow=3)+
      geom_point()

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: I think that 7 is more effective because it is very difficult to see what is happening with each of the ecosystems when they are all plotted on top of each other. 
